---
title: Indicators of social vulnerability explain rates of COVID-19 infections across Canada
author:
  - name: David Beauchesne
    email: david.beauchesne@hotmail.com
    affiliation: uoft
    # correspondingauthor: true
    # footnote: 1
  - name: Eric Liberda
    email: eric.liberda@torontomu.ca
    affiliation: tmu
    #footnote: 2
  - name: Robert Moriarity
    email: rob.moriarity@torontomu.ca
    affiliation: tmu
    #footnote: 2
  - name: Nicholas Spence
    email: nicholas.spence@utoronto.ca
    affiliation: uoft
    #footnote: 2
  - name: Leonard J.S. Tsuji
    email: leonard.tsuji@utoronto.ca
    affiliation: uoft
  - name: Aleksandra Zuk
    email: amz4@queensu.ca
    affiliation: qu
    #footnote: 2
address:
  - code: uoft
    organization: Department of Health and Society, University of Toronto
    addressline: 27 King's College Cir.
    postcode: M5S 1A1
    state: Ontario
    city: Toronto
    country: Canada
  - code: tmu
    organization: 
    addressline: 
    postcode:
    city:
    state: 
    country: 
  - code: qu
    organization: 
    addressline: 
    postcode: 
    city: 
    state: 
    country: 
#footnote:
#  - code: 1
#    text: "This is the first author footnote."
#  - code: 2
#    text: "Another author footnote."
abstract: |
  Write abstract here in yaml header
keywords: 
  - COVID-19
  - Social vulnerability
  - Canada
  - 
  - 
  - 
journal: "Journal of Environmental Management"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: covidfn.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
header-includes:
  - \usepackage{setspace}
  - \usepackage{float}
---

\onehalfspacing

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  echo = FALSE,
  message = FALSE,
  dev = "png",
  dpi = 108,
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  width = 120 # ,
  # table.attr = 'class="flat-table"'
)
```



<!-- ========================================== -->
# Introduction 

The objective of this article was to explore how rates of COVID-19 outcomes during the pandemic were linked to indicators of social vulnerability (*e.g.* employment, income, dwelling condition) across Canada. Our specific goals were to 1) characterize the spatial distribution of social vulnerabilities across Canada, and 2) explore the relationship between social vulnerabilities and rates of COVID-19 infections and deaths during the pandemic. Our goal with this assessment is identify readily available indicators of social vulnerability and target those that should be alleviated to improve our response to another pandemic in the future. 


<!-- ========================================== -->
# Material and methods

## Spatial scope

This assessment is focused on a continental Canada-wide assessment (Figure 1). 


## Data sources

### COVID-19 data 

```{r}
hr <- read.csv("../../data/pipedat/covid_timeline_canada-a56e753b/format/covid_timeline_canada-a56e753b-hr.csv")
covid <- read.csv("../../data/pipedat/covid_timeline_canada-a56e753b/format/covid_timeline_canada-a56e753b-CovidTimelineCanada_hr.csv")
maxDate <- as.Date("2023-05-17")
```

> The Timeline of COVID-19 in Canada (CovidTimelineCanada) is intended to be the definitive source for data regarding the COVID-19 pandemic in Canada. In addition to making available the ready-to-use datasets, this repository also acts as a hub for collaboration on expanding and improving the availability and quality of COVID-19 data in Canada. This repository is maintained by the COVID-19 Canada Open Data Working Group and is one component of the What Happened? COVID-19 in Canada project. 

This dataset provides daily data on COVID-19 infections and deaths for `r nrow(hr)` Canadian health regions throughout the pandemic. We selected data spanning from the onset of the pandemic in Canada (`r min(covid$date)` in the data) until `r maxDate`. 

***Note that the end date should be discussed and decided upon together. Some things to consider for this decision: when certain provinces began phasing out reporting at the level of health regions (see https://github.com/ccodwg/CovidTimelineCanada/blob/main/docs/data_sources/hr_reporting.md), and the advent of rapid testing.***

```{r covid}
dat <- hr |>
  dplyr::group_by(region) |>
  dplyr::summarise(
    region_num = dplyr::n(),
    population = sum(pop, na.rm = TRUE)
  )

tab <- covid |>
  dplyr::filter(date <= maxDate) |>
  dplyr::group_by(name, region) |>
  dplyr::summarize(value_daily = sum(value_daily)) |>
  tidyr::pivot_wider(names_from = name, values_from = value_daily) |>
  dplyr::left_join(dat, by = "region") |>
  dplyr::select(region, region_num, population, cases, deaths)

tab |>
  knitr::kable(
    col.names = c("Province", "Health regions", "Population", "Cases", "Deaths"),
    row.names = FALSE,
    caption = "Number of health regions, population size, and number of cases and deaths for each Canadian province."
  ) # |>
# kableExtra::row_spec(uid, extra_css = "border-bottom: 2px solid")
```


### Indicators of social vulnerability

The whole country was divided into a 1 $km^2$ grid cell to characterize the selected indicators of social vulnerability. The list of indicators used is available in Table 1. The following sections present the different data sources and the indicators they were used to obtain. 

***Note that for a manuscript, I would focus on the indicators of social vulnerability rather than the dataset, but to facilitate our work, I present it here by focusing on datasets.***


```{r indicators}
dat <- data.frame(
  ind = c(
    "Gini index on adjusted household total income",
    "P90/P10 ratio on adjusted household after-tax income",
    "Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)",
    "Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)",
    "Indigenous identity",
    "Children in a one-parent family",
    "Parents in one-parent families",
    "No certificate, diploma or degree",
    "Government transfers (%)",
    "Housing suitability",
    "Dwelling condition",
    "Acceptable housing",
    "Distance to closest road",
    "Distance to closest critical healthcare facility",
    "Distance to closest longterm healthcare facility"
  ),
  src = c(
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census of Population",
    "2021 Census Road Network",
    "Open Database of Healthcare Facilities",
    "Open Database of Healthcare Facilities"
  )
)

dat |>
  knitr::kable(
    col.names = c("Indicators of social vulnerability", "Source"),
    row.names = FALSE,
    caption = "List of indicators of social vulnerability considered for this assessment."
  )
```


#### 2021 Census of Population

Data from the 2021 Census of Population [@statisticscanada2021a] was used to select relevant population indicators as proxies of social vulnerabilities. The indicators were then joined to the census cartographic divisions boundary files for 2021 [@statisticscanada2022f; @statisticscanada2022] and subsequently integrated in the study grid. The selected indicators were:

> - **Gini index on adjusted household total income**: *"The Gini coefficient is a number between zero and one that measures the relative degree of inequality in the distribution of income. The coefficient would register zero (minimum inequality) for a population in which each person received exactly the same adjusted household income and it would register a coefficient of one (maximum inequality) if one person received all the adjusted household income and the rest received none. Even though a single Gini coefficient value has no simple interpretation, comparisons of the level over time or between populations are very straightforward: the higher the coefficient, the higher the inequality of the distribution."*

> - **P90/P10 ratio on adjusted household after-tax income**: *"The P90/P10 ratio is a measure of inequality. It is the ratio of the 90th and the 10th percentile of the adjusted household after-tax income. The 90th percentile means 90% of the population has income that falls below this threshold. The 10th percentile means 10% of the population has income that falls below this threshold."*

> - **Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)**: *"The Low‑income measure, after tax, refers to a fixed percentage (50%) of median adjusted after‑tax income of private households. The household after‑tax income is adjusted by an equivalence scale to take economies of scale into account. This adjustment for different household sizes reflects the fact that a household’s needs increase, but at a decreasing rate, as the number of members increases."*

> - **Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)**: *"The Low‑income cut‑offs, after tax refer to income thresholds, defined using 1992 expenditure data, below which economic families or persons not in economic families would likely have devoted a larger share of their after‑tax income than average to the necessities of food, shelter and clothing. More specifically, the thresholds represented income levels at which these families or persons were expected to spend 20 percentage points or more of their after‑tax income than average on food, shelter and clothing. These thresholds have been adjusted to current dollars using the all‑items Consumer Price Index (CPI)."*

> - **Indigenous identity**: *"Indigenous identity refers to whether the person identified with the Indigenous peoples of Canada. This includes those who identify as First Nations (North American Indian), Métis and/or Inuk (Inuit), and/or those who report being Registered or Treaty Indians (that is, registered under the Indian Act of Canada), and/or those who have membership in a First Nation or Indian band. Aboriginal peoples of Canada (referred to here as Indigenous peoples) are defined in the Constitution Act, 1982, Section 35 (2) as including the Indian, Inuit and Métis peoples of Canada."*

> - **Children in a one-parent family**: *"Percent children living in one-parent family."*

> - **Parents in one-parent families**: *"Percent parent in one-parent family,"*

> - **No certificate, diploma or degree**: *"Percent population with no certificate, diploma or degree, population 25-64 years old."*

> - **Government transfers (%)**: *"Percent of total income composed of government transfers in 2020, corresponding to all cash benefits received from federal, provincial, territorial or municipal governments during the reference period."*


Data from housing suitability [@statisticscanada2022c], dwelling condition [@statisticscanada2022d], and acceptable housing [@statisticscanada2022e] of the 2021 Census of Population [@statisticscanada2021a] were also joined with the 2021 Census cartographic division boundary file [@statisticscanada2022; statisticscanada2022f]. 

> - **Housing suitability**: According to Statistics Canada, housing suitability *"refers to whether a private household is living in suitable accommodations according to the National Occupancy Standard (NOS); that is, whether the dwelling has enough bedrooms for the size and composition of the household. A household is deemed to be living in suitable accommodations if its dwelling has enough bedrooms, as calculated using the NOS. Housing suitability assesses the required number of bedrooms for a household based on the age, sex, and relationships among household members. An alternative variable, persons per room, considers all rooms in a private dwelling and the number of household members. Housing suitability and the National Occupancy Standard (NOS) on which it is based were developed by Canada Mortgage and Housing Corporation (CMHC) through consultations with provincial housing agencies."* Housing suitability was assessed as the proportion of households in a census division considered as not suitable.

> - **Dwelling condition**: Dwelling condition refers to whether the dwelling is in need of repairs. Acceptability of dwelling condition was assessed as the proportion of households in a census division considered as needing major repairs.

> - **Acceptable housing**: According to Statistics Canada, acceptable housing *"refers to whether a household meets each of the three indicator thresholds established by the Canada Mortgage and Housing Corporation for housing adequacy, suitability and affordability. Housing indicator thresholds are defined as follows: 1) adequate housing is reported by their residents as not requiring any major repairs; 2) affordable housing has shelter costs less than 30% of total before-tax household income; 3) suitable housing has enough bedrooms for the size and composition of resident households according to the National Occupancy Standard (NOS), conceived by the Canada Mortgage and Housing Corporation and provincial and territorial representatives. Acceptable housing identifies which thresholds the household falls below, if any. Housing that is adequate in condition, suitable in size and affordable is considered to be acceptable."* Here, acceptable housing was assessed as the proportion of households in a census division that was below any of the thresholds of adequacy, affordability or suitability.


#### Census 2021 road network file

The road network from the Census 2021 road network file [@statisticscanada2021a; statisticscanada2021b] was rasterized as a 1 $km^2$ resolution then integrated in the study grid to obtain a raster of the distribution of the Canadian road network. The distance of the centroid of each cell in the grid to the closest road was then measured to obtain an assessment of the distance to the closest road across Canada.


#### Open Database of Healthcare Facilities

The location of healthcare facilities available in the Open Database of Healthcare Facilities [ODHF; @statisticscanada2020; @statisticscanada2020a] were used to assess the distance to the closest healthcare facility. Facilities were divided between critical and longterm care using the classifications available in the ODHF. Hospitals and ambulatory health care services were considered as critical care facilities, while nursing and residential care facilities were considered as longterm care facilities.


## Assessment

### Integrated dataset 

For each Canadian health unit, we extracted the number of cases, the number of deaths, and the average value of each selected indicator of social vulnerability.

### Correlations

TODO: Assess the correlation between COVID-19 outcomes and indicators of social vulnerability

### Linear regressions

TODO: Build competing regression models to explain COVID-19 outcomes as a function of social vulnerabilities. Perhaps use mixed models and use provinces as a random factor in the assessment

### Cluster analysis

TODO: 

- Identify HR that are most similar in terms of COVID-19 outcomes and social indicators
- Identify variables that explain intra-cluster similarity
- Identify variables that explain inter-cluster dissimilarity


<!-- ========================================== -->
# Results


<!-- ========================================== -->
# Discussion 



<!-- ========================================== -->
# Credit author statememnt  {-}

<!-- https://www.elsevier.com/authors/policies-and-guidelines/credit-author-statement -->

**David Beauchesne**: Conceptualization, Methodology, Software, Formal analysis, Data Curation, Writing - Original Draft, Writing – review & editing, Visualization;
****: 



<!-- ========================================== -->
# Declaration of competing interest  {-}



<!-- ========================================== -->
# Acknowledgements  {-}

<!-- 
List funding sources in this standard way to facilitate compliance to funder's requirements:

Funding: This work was supported by the National Institutes of Health [grant numbers xxxx, yyyy]; the Bill & Melinda Gates Foundation, Seattle, WA [grant number zzzz]; and the United States Institutes of Peace [grant number aaaa].

It is not necessary to include detailed descriptions on the program or type of grants and awards. When funding is from a block grant or other resources available to a university, college, or other research institution, submit the name of the institute or organization that provided the funding.
-->



<!-- ========================================== -->
# Data and code availability {-}

The data used for the cumulative effects assessment of marine shipping in the St. Lawrence and Saguenay Rivers are through Zenodo (DOI: ) and available at the following link: ***add link***. The data shared is the integration of all valued components and stressors within our study grid, which is used to perform the cumulative effects assessment. However, the data shared by First Nations is unavailable due to data sharing agreements. The raw data are not available on Zenodo, as the assessment integrates data from over 80 different projects. However, code is available through a research compendiu (see below) to access all publicly available data, which make up the vast majority of all data used for the assessment. The code used for this assessment is available through a research compendium called *ceanav* and available through a GitHub repository (https://github.com/EffetsCumulatifsNavigation/ceanav) and archived on Zenodo (DOI: ). ***Note:*** *DOIs will be added once review process for publication is completed*

<!--
Scientific results are only as strong as they are reproducible, so being able to identify and use research data is vital to building knowledge. To help make all scientific data more transparent, a number of Elsevier journals encourage their authors to state the availability of their data. With the data statement, authors can be transparent about the data they used in an article and make a statement about its availability together with their published article.
-->

\newpage 

<!-- ========================================== -->
# Figures {-}

<!-- TODO: Update aoi figure to include world map like in the nceastl project -->
<!-- \begin{figure}[H]
\centering
\includegraphics[width=0.95\columnwidth]{../figures/aoi.png}
\caption{Limits of the study area for the cumulative effects assessment of marine shipping activities in the St. Lawrence and the Saguenay rivers in eastern Canada.}
\label{aoi}
\end{figure}
\newpage 
 -->


<!-- ========================================== -->
# Tables {-}


\newpage

# References {-}
